---
title: "Replication Code"
author: "Micah Clark Moody"
output:
  html_document
---

Welcome to this replication code! There are only TWO steps to make sure this runs on your computer.

First, make sure you have all of the library packages installed.

Second, rename the data paths to the data downloaded on your local device. 

# Setup 

```{r}

# path for detention data
dashboard_path <- "YOUR PATH"

# path for reform-oriented data
initiation_path <- "YOUR PATH"

# path for key to clean gun possession data 
gun_level_key_path <- "YOUR PATH"

```

```{r setup}

library(tidyverse)
library(readxl)
library(skimr)
library(lubridate)
library(modelsummary)
library(MASS)               # for ologit models
library(nnet)               # for multinomial logit
library(marginaleffects)    # marginal effect calculations
library(knitr)              # to build kable() table
library(gofcat)             # for brant test
library(tinytable)          # to hold tables in place

```

# Import Data

After this point, the code should run without intervention.

```{r}

# formatting document globally

# rounding decimals at ones place
options(digits = 2)

# setting greyscale for plots
options(ggplot2.discrete.colour = scale_colour_grey, ggplot2.discrete.fill = scale_fill_grey)

```

```{r}

# dashboard offense category vectors

dash_narcotics <- c("Narcotics")
dash_uuw <- c("Unlawful Use of Weapons", "Armed Violence", "Armed Habitual Criminal")
dash_property <- c("Burglary", "Theft", "Retail Theft", "Criminal Damage to Property", "Identity Theft", "Vehicular Invasion", "Arson", "Fraud", "Forgery", "Home Invasion", "Criminal Trespass", "Criminal Trespass to Vehicle")
dash_driving <- c("Driving With Suspended Or Revoked License", "DUI", "Agg DUI", "Failure Report Accident", "Reckless Driving", "Speeding", "Traffic", "PSMV")
dash_arrest_violation <- c("Resisting Police Officer", "Out of County Warrant", "Violation of Order of Protection", "Obstructing Justice", "Escape - Failure to Return", "Crimial Contempt", "Fugitive from Justice", "VOB", "Gun Offender Registration Ordinance", "Fleeing")
dash_violence_against_a_person <- c("Attempt Murder", "Sex Crimes", "Robbery", "Assault", "Battery", "Agg Assault", "Domestic Battery", "Agg Battery", "Murder", "Kidnaping", "Unlawful Restraint", "Criminal Abuse of Elderly", "Reckless Homicide")
dash_other <- c("Not Categorized", "Public Drinking", "Soliciting Unlawful Business", "Reckless Conduct", "Disorderly Conduct", "Prostitution", "Unlawful Use of ID Card", "Public Indecency", "Gambling", "Mob Action", "Terrorism", "NULL", "Reckless Discharge of Firearm", "Agg Discharge of a Firearm", "Hate Crime", "Child Endangerment", "Stalking", "Intimidation", "Harassment")

# initiation offense category vectors

narcotics <- c("Narcotics")
uuw <- c("UUW - Unlawful Use of Weapon", "Gun - Non UUW", "Gun Running")
property <- c("Criminal Damage to Property", "Arson", "Attempt Arson", "Arson and Attempt Arson", "Criminal Trespass To Residence", "Retail Theft", "Residential Burglary", "Burglary", "Theft", "Armed Robbery", "Robbery", "Identity Theft", "Forgery", "Home Invasion", "Credit Card Cases", "Fraudulent ID", "Fraud", "Vehicular Hijacking", "Aggravated Robbery", "Deceptive Practice", "Aggravated Identity Theft", "Theft by Deception", "Attempt Armed Robbery", "Attempt Vehicular Hijacking", "Vehicular Invasion", "Possession Of Burglary Tools")
driving <- c("Aggravated DUI", "DUI", "Driving With Suspended Or Revoked License", "Possession of Stolen Motor Vehicle")
arrest_violation <- c("Escape - Failure to Return", "Failure to Register as a Sex Offender", "Violation Order Of Protection", "Violate Bail Bond", "Possession of Contraband in Penal Institution", "Aggravated Battery Police Officer", "Aggravated Fleeing and Eluding", "Aggravated Assault Police Officer", "Obstructing Justice", "Impersonating Police Officer", "Disarming Police Officer")
violence_against_a_person <- c("Battery", "Attempt Homicide", "Aggravated Battery With A Firearm", "Kidnapping", "Unlawful Restraint", "Homicide", "Child Abduction", "Aggravated Discharge Firearm", "Reckless Homicide", "Domestic Battery")
other <- c("Other Offense", "Reckless Discharge of Firearm", "Prostitution", "Bribery", "Bomb Threat", "Hate Crimes", "Gambling", "Dog Fighting", "PROMIS Conversion", "Possession of Explosives", "Communicating With Witness", "Official Misconduct", "Perjury", "Tampering", "Child Pornography", "Human Trafficking", "Pandering", "Stalking", "Intimidation")

```
```{r}

# IMPORT DASHBOARD DATA

# dashboard returns
dashboard <- read_xlsx(dashboard_path)

# RESTRUCTURE DASHBOARD VARIABLES

# DATE
dashboard$HEARING_DATE <- as.Date(dashboard$HEARING_DATE)

# CATEGORICAL

# detained

dashboard$DETAINED <- ifelse(dashboard$DETAINED == "Yes", 1, 0) #TODO option to explore NULL and unknown to see if 0 is accurate

# NEW VARIABLES

# gun possession charge level and license status

gun_level_key <- read.csv(gun_level_key_path) #TODO option to deselect the redundant columns

dashboard <- dashboard %>%  # merge key with existing data frame setting missing values as NA
    left_join(gun_level_key, by = "TOP_CHARGE") %>% 
    mutate(gun_case = replace_na(gun_case, 0))  #TODO add all gun charges to the key so this statement is accurate. Currently it's all gun charges with 10 or more cases.

# era

# era saving date ranges

foxx_first <- ymd("2016-12-01")
money_bail_last <- ymd("2023-09-17")
no_bail_first <- ymd("2023-09-18")
foxx_last <- ymd("2024-11-30")
burke_first <- ymd("2024-12-01")
burke_last <- ymd("2028-11-30")

# era saving date intervals

money_bail_progressive_sao <- interval(foxx_first, money_bail_last)
no_money_progressive_sao <- interval(no_bail_first, foxx_last)
no_monney_conservative_sao <- interval(burke_first, burke_last)

# era creating new variable

dashboard <- dashboard %>% 
  mutate(era = case_when(HEARING_DATE %within% money_bail_progressive_sao ~ "Era 1", HEARING_DATE %within% no_money_progressive_sao ~ "Reform-Oriented Prosecutor", HEARING_DATE %within% no_monney_conservative_sao ~ "Conservative Prosecutor"))

# era factor variable

dashboard$era <- factor(dashboard$era, levels = c("Reform-Oriented Prosecutor", "Conservative Prosecutor"))

# charge category bigger groups 

dashboard <- dashboard %>% 
  mutate(category = case_when(OFFENSE_CATEGORY %in% dash_narcotics ~ "narcotics", OFFENSE_CATEGORY %in% dash_uuw ~ "uuw", OFFENSE_CATEGORY %in% dash_property ~ "property", OFFENSE_CATEGORY %in% dash_driving ~ "driving", OFFENSE_CATEGORY %in% dash_arrest_violation ~ "arrest_violation", OFFENSE_CATEGORY %in% dash_other ~ "other", OFFENSE_CATEGORY %in% dash_violence_against_a_person ~ "violence_against_a_person"))

dashboard$category <- factor(dashboard$category, levels = c("arrest_violation", "driving", "narcotics", "other", "property", "uuw", "violence_against_a_person"), labels = c("Arrest and Violation", "Driving", "Narcotics", "Other", "Property", "Gun Possession", "Violence Against a Person"))

```

```{r}

# IMPORT INITIATION DATA

# initiation from open data portal
initiation_all_charges <- read.csv(initiation_path)
initiation <- filter(initiation_all_charges, PRIMARY_CHARGE_FLAG == "true")

# RESTRUCTURE INITIATION DATA

# DATE VARIABLES

# initiation open data 
initiation$RECEIVED_DATE <- as_date(initiation$RECEIVED_DATE)

# CATEGORICAL VARIABLES

# race

recode_race <- function(x){case_when(x %in% c("Black") ~ "black", x %in% c("HISPANIC", "White [Hispanic or Latino]", "White/Black [Hispanic or Latino]") ~ "latinx", x %in% c("White", "CAUCASIAN") ~ "white", x %in% c("American Indian", "Native Hawaiian or Other Pacific Islander") ~ "native", x %in% c("Asian", "ASIAN") ~ "asian", x %in% c("Biracial", "Middle Eastern/North African", "Albino", "", "Unknown") ~ "unclear")}

race_fct_lvl <- c("white", "black", "latinx", "native", "asian", "unclear")

initiation$RACE <- recode_race(initiation$RACE)

initiation$RACE <- factor(initiation$RACE, levels = race_fct_lvl, ordered = FALSE)

# charge level

initiation <- initiation %>% 
    filter(CLASS %in% c("X", "2", "3", "4"))

initiation$CLASS <- factor(initiation$CLASS, levels = c("4", "3", "2", "X"), ordered = TRUE)

# bond type initial

initiation$BOND_TYPE_INITIAL <- factor(initiation$BOND_TYPE_INITIAL, levels = c("I Bond", "D Bond", "C Bond", "No Bond"), ordered = TRUE)

# CREATE VARIABLES

# offense category

initiation <- initiation %>%
    mutate("category" = case_when(UPDATED_OFFENSE_CATEGORY %in% narcotics ~ "narcotics", UPDATED_OFFENSE_CATEGORY %in% uuw ~ "uuw", UPDATED_OFFENSE_CATEGORY %in% property ~ "property", UPDATED_OFFENSE_CATEGORY %in% driving ~ "driving", UPDATED_OFFENSE_CATEGORY %in% arrest_violation ~ "arrest_violation", UPDATED_OFFENSE_CATEGORY %in% other ~ "other", UPDATED_OFFENSE_CATEGORY %in% violence_against_a_person ~ "violence_against_a_person"))

# era (if needed)

# NUMERIC 

initiation$BOND_AMOUNT_INITIAL <- as.numeric(gsub(",", "", initiation$BOND_AMOUNT_INITIAL))

```

```{r}

era <- c("Era 1", "Era 2", "Era 3")
prosecutor <- c("Reform-Oriented", "Reform-Oriented", "Conservative")
bail <- c("Money Bail", "No Money Bail", "No Money Bail")
dates <- c("Dec. 1, 2018 to Sept. 17, 2023", "Sept. 18, 2023 to Nove. 31, 2024", "December 1, 2024 to Present")
dataset <- c("Reform-Oriented Only", "Reform-Oriented and Detention", "Detention Only")
unique <- c("Race, Gender, Class", "Race, Gender, Class and Prosecutor Detention Request, Detention", "Prosecutor Detention Request, Detention")

dataset_table <- data.frame(era, prosecutor, bail, dates, dataset, unique)

tt(dataset_table, caption = "Era and Dataset Description", col.names = c("Era", "Dates", "Bail Law", "Prosecutor", "Dataset", "Unique Var."), placement = "H", width = c(0.1, 0.1, 0.1, 0.2, 0.1, 0.4))

```

# Hypothesis 1: Change in gun possession detention is driving detention rates. Supported. 

```{r}

# run models

logistic_detained_era <-  glm(DETAINED ~ relevel(era, ref = "Reform-Oriented Prosecutor")*category, family = binomial, data = dashboard)

lpm_detained_era <- lm(DETAINED ~ relevel(era, ref = "Reform-Oriented Prosecutor")*category, data = dashboard)

probit_detained_era <- glm(DETAINED ~ relevel(era, ref = "Reform-Oriented Prosecutor")*category, family = binomial(probit), data = dashboard)

# identical AIC and BIC
# don't present together because logit is in log-odds or odds ratio while probit is in change in z-score

```

```{r}

plot_predictions(lpm_detained_era, condition = c("category", "era"), vcov = "HC3") + labs(title = "Linear probabiltiy model predicted probabiltiy of detention for era and charge category", x = "Charge Category", y = "Predicted Probability of Detention", color = "Prosecutor Era") + theme_minimal() + theme(legend.position = "bottom")

```

## Hypothesis 2: There are lower detention rates for gun possession with a license. Supported.

```{r}

# create dataset

gun_data <- dashboard %>% 
  filter(gun_case == 1) %>% 
  filter(top_charge_level == "level_4") %>% 
  filter(gun_license %in% c("No FOID or CCL", "FOID no CCL", "FOID and CCL"))

gun_data$gun_license <- factor(gun_data$gun_license, levels = c("No FOID or CCL", "FOID no CCL", "FOID and CCL", "Unclear", ""), ordered = FALSE)

```

```{r}

# generate models

binom_era <- glm(DETAINED ~ era, family = binomial, data = gun_data)

binom_era_lic_level <- glm(DETAINED ~ era + relevel(gun_license, ref = "No FOID or CCL"), family = binomial, data = gun_data)

```

```{r}

coef_list_2 <- c("eraConservative Prosecutor" = "Conservative Prosecutor", "relevel(gun_license, ref = \"No FOID or CCL\")FOID and CCL" = "FOID and CCL", "relevel(gun_license, ref = \"No FOID or CCL\")FOID no CCL" = "FOID, no CCL")

# present models

hyp_2_model <- modelsummary(list("Era Only" = binom_era, "Era and License" = binom_era_lic_level), exp = TRUE, ci_method = "wald", fmt = 3, stars = TRUE, gof_map = c("nobs", "aic", "bic"), coef_map = coef_list_2, title = "Odds Ratio Results from Logit Model for Gun Possession Charges", output = "tinytable")

theme_latex(hyp_2_model, placement = "H")

era_license_count <- gun_data %>% 
  group_by(era, gun_license) %>% 
  summarise(n())

```

## Hypothesis 3: Black defendants receive harsher release types than non-Black people.

```{r}

# create dataset

initiation_demo_filter <- initiation %>% 
    filter(GENDER %in% c("Female", "Male")) %>% 
    filter(RACE %in% c("black", "white", "latinx")) %>% 
    filter(!is.na(category)) %>% 
    filter(!is.na(AGE_AT_INCIDENT))

initiation_demo_filter$RACE <- factor(initiation_demo_filter$RACE, levels = c("black", "white", "latinx"))

```

```{r}

# create models

model_race <- polr(BOND_TYPE_INITIAL ~ RACE, Hess = T, data = initiation_demo_filter)

model_demo <- polr(BOND_TYPE_INITIAL ~ RACE + GENDER + AGE_AT_INCIDENT, Hess = T, data = initiation_demo_filter)

model_class_category <- polr(BOND_TYPE_INITIAL ~ CLASS + category, Hess = T, data = initiation_demo_filter)

model_race_gender_class_category <- polr(BOND_TYPE_INITIAL ~ RACE + GENDER + CLASS + category, Hess = T, data = initiation_demo_filter)

model_demo_class_category <- polr(BOND_TYPE_INITIAL ~ RACE + GENDER + AGE_AT_INCIDENT + CLASS + category, Hess = T, data = initiation_demo_filter)

```

```{r}

# display models

coef_subset <- c("I Bond|D Bond" = "I Bond|D Bond", "D Bond|C Bond" = "D Bond|C Bond", "C Bond|No Bond" = "C Bond|No Bond", "RACEwhite" = "White", "RACElatinx" = "Latinx")

hyp_3a_model <- modelsummary(list("Race Only" = model_race, "Charge Only" = model_class_category, "Demographics and Charge" = model_demo_class_category), exp = TRUE, ci_method = "wald", fmt = 3, stars = TRUE, coef_map = coef_subset, title = "Ordered Logistic Regression Odds Ratio Models with Race, Charge, and Controls on Bond Type", output = "tinytable")

theme_latex(hyp_3a_model, placement = "H")

```

## Hypothesis 4: The defendant being black is correlated with a higher bond amount. Supported.

```{r}

# create dataset for only cases where money bail is set

initiation_bond_set <- initiation_demo_filter %>%  # starting with the same dataset as the analysis with demographic controls
    filter(BOND_TYPE_INITIAL %in% c("D Bond", "C Bond")) %>%

    # create a column of release cost (10% of D Bond amount)
    mutate(release_cost = ifelse(BOND_TYPE_INITIAL == "D Bond", BOND_AMOUNT_INITIAL/10, BOND_AMOUNT_INITIAL)) %>% 
    filter(!is.na(release_cost), release_cost > 0)


initiation_bond_set$CLASS <- factor(initiation_bond_set$CLASS, levels = c("4", "3", "2", "X"), ordered = FALSE)

# test for skew by comparing mean and median

mean_release_cost <- mean(initiation_bond_set$release_cost)
median_release_cost <- median(initiation_bond_set$release_cost)

```

```{r}

# create models

lm_race_bond <- lm(log(release_cost) ~ factor(RACE, ordered = FALSE) + AGE_AT_INCIDENT + GENDER, data = initiation_bond_set)

lm_class_bond <- lm(log(release_cost) ~ relevel(CLASS, ref = "4") + category, data = initiation_bond_set)

lm_race_class_bond <- lm(log(release_cost) ~ factor(RACE, ordered = FALSE) + AGE_AT_INCIDENT + GENDER + relevel(CLASS, ref = "4") + category, data = initiation_bond_set)

```

```{r}

# present results

coef_list <- c("(Intercept)" = "Intercept", "factor(RACE, ordered = FALSE)black" = "Black", 'relevel(factor(RACE, ordered = FALSE), ref = "black")white' = "White", "factor(RACE, ordered = FALSE)latinx" = "Latinx", 'relevel(factor(RACE, ordered = FALSE), ref = "black")latinx' = "Latinx", "factor(RACE, ordered = FALSE)native" = "Native American", 'relevel(factor(RACE, ordered = FALSE), ref = "black")native' = "Native American", "factor(RACE, ordered = FALSE)asian" = "Asian", 'relevel(factor(RACE, ordered = FALSE), ref = "black")asian' = "Asian", "factor(RACE, ordered = FALSE)unclear" = "Race Unclear", 'relevel(factor(RACE, ordered = FALSE), ref = "black")unclear' = "Race Unclear", "AGE_AT_INCIDENT" = "Age", "GENDERMale" = "Male", 'relevel(CLASS, ref = "4")3' = "Class 3", 'relevel(CLASS, ref = "4")2' = "Class 2", 'relevel(CLASS, ref = "4")X' = "Class X")

hyp_3b_model <- modelsummary(list("Demographics Only" = lm_race_bond, "Charge Only" = lm_class_bond, "Demographics and Charge" = lm_race_class_bond), fmt = 3, gof_map = c("nobs", "aic", "bic"), stars = TRUE, coef_map = coef_list, title = "Table Comparing Linear Regression Results for Race and Charge on Log Money Bail Amount", subtitle = "Charge category results are calculated but no presented. See Appendix for all coefficients.", output = "tinytable")

theme_latex(hyp_3b_model, placement = "H")

```

# Conclusion

```{r}

hyp <- c("Hypothesis 1", "Hypothesis 2", "Hypothesis 3", "Hypothesis 4")
hyp_words <- c("Change in gun possession detention is driving detention rates.","There are lower detention rates for gun possession with a license.", "Black defendants receive harsher release types than non-Black people.", "The defendant being black is correlated with a higher bond amount.")
hyp_outcome <- c("Supported", "Supported", "Mixed Results", "Supported")

hyp_table <- data.frame(hyp, hyp_words, hyp_outcome)

colnames(hyp_table) <- c("Number", "Hypothesis", "Finding")

hyp_table <- tt(hyp_table, caption = "Hypothesis Findings")

theme_latex(hyp_table, placement = "H")

```

# Appendix

## Hypothesis 1: Change in gun possession detention is driving detention rates.

```{r}

coef_list_1_app <- c("relevel(era, ref = \"Reform-Oriented Prosecutor\")Conservative Prosecutor" = "Conservative Prosecutor", "categoryDriving" = "Driving", "categoryNarcotics" = "Narcotics", "categoryOther" = "Other", "categoryProperty" = "Property", "categoryGun Possession" = "Gun Possession", "categoryViolence Against a Person" = "Violence Against a Person", "relevel(era, ref = \"Reform-Oriented Prosecutor\")Conservative Prosecutor:categoryDriving" = "Conservative × Driving", "relevel(era, ref = \"Reform-Oriented Prosecutor\")Conservative Prosecutor:categoryNarcotics" = "Conservative × Narcotics", "relevel(era, ref = \"Reform-Oriented Prosecutor\")Conservative Prosecutor:categoryOther" = "Conservative × Other", "relevel(era, ref = \"Reform-Oriented Prosecutor\")Conservative Prosecutor:categoryProperty" = "Conservative × Property", "relevel(era, ref = \"Reform-Oriented Prosecutor\")Conservative Prosecutor:categoryGun Possession" = "Conservative × Gun Possession", "relevel(era, ref = \"Reform-Oriented Prosecutor\")Conservative Prosecutor:categoryViolence Against a Person" = "Conservative × Violence Against a Person")

app_1 <- modelsummary(list("Logistic Log-Odds" = logistic_detained_era, "Probit Change in Z-Score" = probit_detained_era, "LPM Model" = lpm_detained_era), ci_method = "wald", fmt = 3, stars = TRUE, gof_map = c("nobs", "aic", "bic"), coef_map = coef_list_1_app, output = "tinytable")

theme_latex(app_1, placement = "H")

```

## Hypthesis 3: Race on Bond Type Model Comparison 

```{r display-models}

coef_list <- c("I Bond|D Bond" = "I Bond|D Bond", "D Bond|C Bond" = "D Bond|C Bond", "C Bond|No Bond" = "C Bond|No Bond", "RACEwhite" = "White", "RACElatinx" = "Latinx", "GENDERMale" = "Male", "AGE_AT_INCIDENT" = "Age", "CLASS.L" = "Class X", "CLASS.Q" = "Class 2", "CLASS.C" = "Class 3", "categorydriving" = "Driving", "categorynarcotics" = "Narcotics", "categoryother" = "Other", "categoryproperty" = "Property", "categoryuuw" = "Gun Possession", "categoryviolence_against_a_person" = "Violence Against Another Person")

hyp_3_app <- modelsummary(list("Race Only" = model_race, "Demographics and Charge" = model_demo_class_category), ci_method = "wald", fmt = 3, stars = TRUE, gof_map = c("nobs"), coef_map = coef_list, output = "tinytable")

theme_latex(hyp_3_app, inner = "cells={font=\\small}", placement = "H")

```

```{r}

# brant test model

hyp_3_brant_results <- brant.test(model_race_gender_class_category)

```

## Hypothesis 4: Race and Charge on Money Bail Amount Comparison

```{r}

# present results

coef_list <- c("(Intercept)" = "Intercept", "factor(RACE, ordered = FALSE)black" = "Black", 'relevel(factor(RACE, ordered = FALSE), ref = "black")white' = "White", "factor(RACE, ordered = FALSE)latinx" = "Latinx", 'relevel(factor(RACE, ordered = FALSE), ref = "black")latinx' = "Latinx", "AGE_AT_INCIDENT" = "Age", "GENDERMale" = "Male", 'relevel(CLASS, ref = "4")3' = "Class 3", 'relevel(CLASS, ref = "4")2' = "Class 2", 'relevel(CLASS, ref = "4")X' = "Class X", "categorydriving" = "Driving", "categorynarcotics" = "Narcotics", "categoryother" = "Other", "categoryproperty" = "Property", "categoryuuw" = "Gun Possession", "categoryviolence_against_a_person" = "Violence Against Another Person")

hyp_3b_app <- modelsummary(list("Demographics Only" = lm_race_bond, "Charge Only" = lm_class_bond, "Demographics and Charge" = lm_race_class_bond), fmt = 3, gof_map = c("nobs", "aic", "bic"), stars = TRUE, coef_map = coef_list, title = "Table Comparing Linear Regression Results for Race and Charge on Money Bail Amount", subtitle = "Charge category results are calculated but no presented. See Appendix for all coefficients.", output = "tinytable")

theme_latex(hyp_3b_app, placement = "H")

```